Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-07-16 paper.Rnw
Rest of Mallar\'s comments for methods/results

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Winterburn J., Lett, T., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\maketitle

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007,Weiner2013}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

@@

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002},
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
Yushkevich and colleagues \citep{Yushkevich2009} acquired and labelled
hippocampal subfields on high-resolution MRI data from post-mortem medial
temporal lobe samples an demonstrate it's applicability in the segmentation of
other MRI-volumes using nonlinear registration guided using manually derived
hippocampus masks and specific landmarks. In their work they demonstrate
accurate parcellations of the subfields.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.
The optimal parameter settings found in Experiment 1 are used.

@@

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject T1 BRAVO
images (referred to as the {\em BRAVO} dataset), and the five
Winterburn atlas subject images subsampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em Subsampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used.

\paragraph{Evaluation method}  
\mb performance is measured with respect to straight-forward nearest-neighbour
resampling of the Winterburn atlas labels to 0.9mm-isotropic voxels (i.e. no 
label propagation). The volumes of each hippocampal subregion in the subsampled
Winterburn atlas labels are compared to the volumes obtained from the full 
resolution Winterburn atlas labels, the \mb labels on the subsampled Winterburn 
atlas images, and the \mb labels on the T1 BRAVO acquisitions.

@@

The hippocampus were traced following a validated protocol developed by Dr Jens
Pruessner \todo{cite: (Pruessner et al., 2000)}. A recent update to this
protocol by Dr J Pruessner in 2006 allows to accurately and consistently
subdivide the hippocampus into three different subregions: head, body, and tail.
We use these segmentations as to validate our implementation of \mb.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the SZ-FEP dataset are segmented.  The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.

@@

\todo{done elsewhere?}

@@

\paragraph{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct evaluation between labels
produced by \mb and the ADNI labels in terms of overlap; we would not expect
different segmentation protocols to have a high degree of overlap.  Instead, to
evaluate the performance of \mb we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes. Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: show variability of MAGeT over parameters
\begin{figure}
<<ADN1-Xval-variability, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ANTSmajvote = subset(all_data, 
                     reg_method=="ANTS" & 
                     method.mb=="Majority Vote" &
                    templates.mb %in% seq(1,20,2))

data.stats = ddply(ANTSmajvote, c("subject", "label","atlases","templates.mb"), function (df) {
  data.frame(
    MA.sd  =  sd(df$k.ma),
    MB.sd  =  sd(df$k.mb),
    MA.var = var(df$k.ma),
    MB.var = var(df$k.mb)
  )
})

@@

ggplot(subset(stats.tests, direction > 0, stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation 
  is computed for each subject across all ten rounds of validation. Shown on the 
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test 
  comparing the distribution of variances at each parameter setting 
  (atlas/template library size). Only points where \mb mean variability is lower 
  than multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentations of hippocampal subfields. To
achieve this, a leave-one-out validation is conducted in which lower-resolution
images ($0.9mm^3$ voxels) of each Winterburn atlas subject is segmented using 
the remaining Winterburn atlases. As a point of comparison, volumes of Winterburn 
atlases when subsampled to $0.9mm^3$ voxels are also computed.

In general, across hippocampal subregions the percent error in volume between
\mb segmentations and the manual Winterburn atlas segmentations compares
favourably to error when resampling the atlas segmentations (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate
subregions all show near or smaller percent errors. The Sibiculum and CA2/CA3
subregions show distinctly larger volumes than resampling error. 
\todo{note direction of change}.

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_boxplot() +
  labs(y="Percent error in volume", x="Region") +
  scale_colour_discrete("") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Percent error in segmentation volume by hippocampus subregion.} 
  Percent error is measured against the volumes of of the unmodified Winterburn
  atlas segmentations.
  {\bf Subsampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + WA Subsampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + WA BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the MAGeT brain labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
\end{figure}

@@

\mb produces hippocampus segmentation volumes that are highly correlated with 
manual segmentation volumes (Figure \ref{fig:SZ-volumes}). 
\todo{p-value}

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup">>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
casted = dcast(SZ_volumes, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Manual unilateral whole hippocampus volume " (mm^3))) + 
  ylab(expression("MAGeT unilateral whole hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

@@

\begin{figure}[h]
<<ADNI-scr-volumes-boxplot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=7,fig.height=4>>=
melted = melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT"),
              variable.name="Method", value.name="Volume")
correlations = ddply(melted, c("DX","Method"), function (df) {
  data.frame(
    pearson = cor(df$SNT, df$Volume)
  )
})

qplot(DX,value,data=melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT")),
      colour=variable,geom="boxplot") + 
      geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
                colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Bilateral whole hippocampal volume (", mm^3, ")"))) + 
      theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT) by disease category.  Pearson correlation 
  with manual volumes are shown below each box-and-whisker.}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

@@

\begin{figure}
<<ADN1-src-Bland-Atlman, cache=TRUE, dependson='ADNI-baseline-volumes-prep', fig.width=6, fig.height=6>>=
melted=melt(totals.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")
melted$diff = melted$SNT - melted$volume
melted$mean = ( melted$SNT + melted$volume ) / 2
limits = ddply(melted, c("method"), function (df) { 
  data.frame(
    y = mean(df$diff) + c(-1.96,0,2) * sd(df$diff) 
)})

ggplot(melted, aes(x= mean, y = diff, colour=DX)) + 
  facet_wrap( ~ method, nrow=2, scales="free") + 
  geom_point() +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown", data=limits) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits, hjust=1.1,vjust=-0.8, size=3) + 
  xlab('Mean of manual and automated total hippocampal volume') + 
  ylab('Difference between manual and automated total hippocampal volume') + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  geom_smooth(method="lm", formula = y~x) + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Bland-Altman plot comparing manual (SNT) and automated total hippocampal volumes 
  in the \adnidataset dataset.  The overall mean difference, and limits of 
  agreement ($\pm 1.96SD$) are shown by dashed horizontal lines.}
  \label{fig:ADNI-scr-Bland-Altman}
\end{figure}

@@

>>>>

\title{Bootstrapping Multi-atlas Hippocampal Segmentation with \mb}
\author{Pipitone J., Park M., Winterburn J., Lett, T., Lerch J., Pruessner J., Lepage M., \\ 
Voineskos A., Chakravarty M.M., and \\ 
the Alzheimer's Disease Neuroimaging Initiative}
\todo{Full names}
\todo{Associations}
\maketitle

@@

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}.  In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}.  In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater. However
the rapid increase in the availability of MRI data and the time and expertise
required for manual segmentation is prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007,Weiner2013}.
Further, there is little agreement between researchers regarding how exactly the
hippocampus should be identified in MRI images \citep{Geuze2004} and this has
led to efforts to create an unified hippocampal segmentation protocol
\citep{Jack2011}.

@@

One methodology that can be used to mitigate these sources of errors involves
the use of multiple manually segmented atlases and probabilistic segmentation
techniques, such as those found in the FreeSurfer package \citep{Fischl2002}.
FreeSurfer uses a probabilistic atlas of anatomical and tissue classes along
with spatial constraints for class labels encoded using a Markov random field
model to segment the entire brain.

@@

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampal subfields using standard T1-weighted data.
The first is included with the FreeSurfer package \citep{VanLeemput2009}.  This
work is limited as it omits the tail of the hippocampus and the segmentation
protocol has yet to be fully validated.  Nonetheless, they demonstrate that the
applicability of their work using data from 10 subjects.  In the second method,
Yushkevich and colleagues \citep{Yushkevich2009} acquired and labelled
hippocampal subfields on high-resolution MRI data from post-mortem medial
temporal lobe samples and demonstrate accurate parcellations of hippocampal subfields in
other MRI-volumes using nonlinear registration guided using manually derived
hippocampus masks and specific landmarks. \todo{flow is terrible in this sentence...}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Hippocampal Subfield Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design. 
The optimal parameter settings for \mb found in Experiment 1 are used in this 
experiment.

@@

In this experiment the Winterburn atlases are used as the cross-validation dataset. 
The five $300\mu$-isotropic voxel images and labels are used as atlases, and 
LOOCV is conducted using the five Winterburn atlas subject images 
subsampled (using trilinear interpolation) to 0.9mm-isotropic voxel 
resolution (referred to as the {\em Subsampled} dataset) as input subjects. Subsampling
the subject images allows us to assess \mb in a typical segmentation scenario (high-resolution atlases
and lower-resolution subjects).  The template library is composed of the subject
images, plus an additional set of 3T T1 images (0.9mm-isotropic voxels) of healthy
subjects acquired separately (Table \ref{table:WAval-healthy-demographics}). The 
optimal size of template library, registration method, and label fusion method 
found in Experiment 1 are used. Each resampled Winterburn atlas subject image is 
segmented by \mb with that subject's image excluded from the atlas library.

We reproduce this experiment in a separate LOOCV run in which the resampled 
Winterburn atlas subject images are substituted for separately acquired T1 
BRAVO images of four of the subjects (referred to as the {\em BRAVO} dataset).

\paragraph{Evaluation method}  
To assess the \mb LOOCV segmentations we compute the relative percent error in 
hippocampal volume with the full resolution Winterburn atlas 
segmentations. In addition, by computing the relative error in volume of
the Winterburn atlas labels resampled (with nearest-neighbour interpolation) to 0.9mm-isotropic voxels, we obtain a baseline error to assess against.

@@

Each subject hippocampus is traced following a validated segmentation protocol 
\citep{Pruessner2000}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the SZ-FEP dataset are segmented by \mb.  The 
optimal size of template library, registration method, and label fusion method 
found in Experiment 1 are used.

@@

\todo{take description from below}

@@

\paragraph{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for evaluation of by measuring overlap. 
That is, different segmentation protocols will necessarily produce segmentations
that do not perfectly overlap, and so the degree of overlap cannot be used soley 
to assess two segmentaiton methods that each use different protocols.

Instead, we assess the correlation in hippocampal volume between \mb and manual segmentation
(SNT). For reference, we also compute the correlation in hippocampal volume between
the existing, established automated segmentation methods: FSL, FreeSurfer, and 
MAPER.  Additionally, we evaluate the volume-related fixed and proportional 
biases in all segmentation methods using Bland-Altman plots \todo{Bland-Altman ref}.

\todo{references} with manual segmentations. 
\todo{"hippocampal volume" = mean hemispheric volume?}

@@

\begin{figure}

<<ADN1-xval-k-mean-mb.ma, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ss = subset(all_data_mean, reg_method == 'ANTS' & method.mb == 'Majority Vote')
ss = subset(ss, (templates.mb*atlases) %% 2 == 1)
ma_lines = ddply(ss, c("atlases"), function (df) {data.frame(yint = mean(df$k.ma))})

ggplot(ss,
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,geom='point',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  geom_hline(yintercept=ma_lines$yint, colour = c("red", "orange", "green", "cyan", "purple"), linetype="dashed") + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \label{fig:ADNI1-xval-k_mean}
  \caption{
  {\bf Mean DSC of \mb segmentations across 69 ADNI1 subjects, by atlas and
  template library size, registration algorithm, and label fusion method.} 
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: show variability of MAGeT over parameters
\begin{figure}
<<ADN1-Xval-variability, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
ants = subset(all_data_mean, 
                     reg_method=="ANTS" & 
                     method.mb=="Majority Vote" &
                    templates.mb %in% seq(1,20,2))

ants.stats = ddply(ants, c("subject", "atlases","templates.mb"), 
                   function (df) {
                     data.frame(
                       MA.sd  =  sd(df$k.ma),
                       MB.sd  =  sd(df$k.mb),
                       MA.var = var(df$k.ma),
                       MB.var = var(df$k.mb)
                       )
                   })

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation 
  is computed for each subject across all ten rounds of validation. Shown on the 
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test 
  comparing the distribution of variances at each parameter setting 
  (atlas/template library size). Only points where \mb mean variability is lower 
  than multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
This experiment explores \mb segmentation of hippocampal subfields. To
achieve this, a leave-one-out validation is conducted in which lower-resolution
images ($0.9mm^3$ voxels) of each Winterburn atlas subject are segmented using 
the remaining Winterburn atlases.

In general, across hippocampal subregions the percent error in volume of
\mb segmentations compares
favourably to the error resulting from image resampling (Figure
\ref{fig:WAval-vol-boxplot}). In particular, the CA1, CA4, and Dentate
subregions all show near or smaller percent errors. The Sibiculum and CA2/CA3
subregions show distinctly larger error than is found through resampling.  
\todo{note direction of change}.

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() +
  labs(y=expression(paste("Percent error in volume relative to 0.3", mm^3, " Winterburn Atlases")), x="Region") +
  scale_colour_discrete("") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Percent error in segmentation volume by hippocampus subregion.} 
  Percent error is measured against the volumes of of the unmodified Winterburn
  atlas segmentations.
  {\bf Subsampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + WA Subsampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + WA BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/subfield-montage.pdf}
  \end{centering}
  \caption{Sagittal slices from two subjects showing a comparison of the
  original Winterburn atlas subfield segmentations (at 0.3mm3-isotropic voxel
  resolution), the subsampled Winterburn segmentations (at 0.9mm3-isotropic
  voxel resolution), and the MAGeT brain labels on the subsampled atlas image.}
  \label{fig:subfield-montage}
  \todo{MAGeT on BRAVO.  Also, show coronal views}
\end{figure}

@@

\mb produces hippocampus segmentation volumes that are highly correlated with 
manual segmentation volumes (Figure \ref{fig:SZ-volumes}). 
\todo{r and p-values}

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup">>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
mean_sz_vols = ddply(SZ_volumes, c('Subject', 'Method'), function (df) {
  data.frame(Volume = mean(df$Volume))})

casted = dcast(mean_sz_vols, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Mean Manual hippocampus volume " (mm^3))) + 
  ylab(expression("Mean MAGeT hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volume for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}

@@

\begin{figure}[h]
<<ADNI-scr-volumes-boxplot,dependson='ADNI-baseline-volumes-prep',cache=T,fig.width=7,fig.height=4>>=
melted = melt(totals.complete,measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"),
              variable.name="Method", value.name="Volume")
correlations = ddply(melted, c("DX","Method"), function (df) {
  data.frame(
    pearson = cor(df$SNT, df$Volume)
  )
})
m2 = melt(totals.complete,measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT","SNT"), variable.name="Method")
levels(m2$Method) = c("FreeSurfer"= "FreeSurfer", "FSL" = "FSL", "MAPER" = "MAPER", "MAGeT" = "MAGeT", "SNT" = "Manual")
m2$DX = factor(m2$DX, levels(m2$DX)[c(2,3,1)])

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Combined whole hippocampal volume (", mm^3, ")"))) + 
      theme(legend.direction = "horizontal", legend.position = "bottom", text=element_text(size=18))
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT) by disease category.  Pearson correlation 
  with manual volumes are shown below each box-and-whisker.}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

@@

\begin{figure}
<<ADN1-src-Bland-Atlman, cache=TRUE, dependson='ADNI-baseline-volumes-prep', fig.width=6, fig.height=6>>=
names(totals.complete)[8] = "FreeSurfer"
melted=melt(totals.complete, measure.vars=c("FreeSurfer", "FSL", "MAPER","MAGeT"), 
            variable.name = "method", value.name = "volume")

melted$diff = melted$SNT - melted$volume
melted = subset(melted, (method == "FreeSurfer") | 
                        (method == "FSL"   & diff > -5000 & diff < 900) | 
                        (method == "MAPER" & diff > -500 & diff < 1000) | 
                        (method == "MAGeT" & diff > -1100))
#melted = subset(melted, method == "MAGeT")
melted$DX = factor(melted$DX, levels(melted$DX)[c(2,3,1)])
melted$mean = ( melted$SNT + melted$volume ) / 2
limits = ddply(melted, c("method"), function (df) { 
  data.frame(
    y = mean(df$diff) + c(-1.96,0,2) * sd(df$diff) 
)})

names(melted)[4] = "Diagnosis"
ggplot(melted, aes(x= mean, y = diff, colour=Diagnosis)) + 
  facet_wrap( ~ method, nrow=2, scales="free") + 
  geom_point() +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown", data=limits) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits, hjust=1.1,vjust=-0.8, size=3) + 
  xlab(expression('Mean of manual and automated volume' (mm^3))) + 
  ylab(expression('Difference in manual and automated volumes' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  geom_smooth(method="lm", formula = y~x) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", text=element_text(size=20))
@
  \caption{Bland-Altman plot comparing manual (SNT) and automated total hippocampal volumes 
  in the \adnidataset dataset.  The overall mean difference, and limits of 
  agreement ($\pm 1.96SD$) are shown by dashed horizontal lines.}
  \label{fig:ADNI-scr-Bland-Altman}
\end{figure}

@@

In this manuscript we have presented the implementation and validation of the MAGeT Brain framework; a methodology that requires very few input atlases in order to provide accurate and reliable segmentations.  We have demonstrated that our methodology robustly provides accurate and consistent segmentations in populations with different ageing and neuropsychiatric characteristics.  Further, we have demonstrated that algorithmic performance is not dependent on a single definition of the hippocampus and can easily differing hippocampal definitions \cite{Winterburn,Pruessner,and SNT}.  Finally, we also demonstrate that MAGeT Brain provides accurate automatic identification of the hippocampal subfields despite contrast and resolution limitations in standard T1-weighted image volumes.

Throughout the cross validation in Experiment #1 (10-fold cross validation in the ADNI sample) we demonstrate that two parameter choices increase segmentation accuracy; namely increasing the number of atlases and the number of templates.  However after setting the parameters to 5 atlases and 15 templates that there are diminishing returns with respect to this improvement.  We were somewhat surprised to find that cross-correlation and normalized mutual information based weighted voting did little to improve segmentation accuracy.  This suggests that using an intermediate template library generated by accurate nonlinear registration methods such as ANTs \cite{Avants|} sufficiently filters sources of error such as nonlinear registration error and partial volume effects that arise through the use of nearest neighbour resampling. From our previous work on the MAGeT Brain algorithm we have shown that these increases in error are not simply a smoothing or averaging effect \cite{Chakravarty}.  Although, the goal of this manuscript was to not exhaustively test or validate multiple different voting strategies in the context of our segmentation algorithm, it is important to note that other strategies for voting are available.  For example, other groups have used the STAPLE algorithm \cite{Warfield} (or variants of the STAPLE algorithm \cite{stuff from the Duchesne group here}).  Can then be weighted each segmentation based upon its estimated performance level with respect to the other available candidate segmentations.  Further, the sensitivity and specificity parameters could also be tuned to potentially improve segmentation accuracy and reliability.  It is rather likely that using more sophisticated voting methods would have a positive effect on the overall segmentation performance.  To this end, more work is required in order to determine the source of the slight decrease in segmentation performance when the number of templates are set to an even number.  Our initial concern was that this dip in performance was a by-product of the MAGeT Brain algorithm itself;  however we determined that this pattern was also true in the analysis of multi-atlas segmentations we used in our experiments.  We believe that our voting methodology is biased towards labels with the lowest numeric values when breaking ties, thus causing the slight bias observed when using an even number of templates.  This is another area where the voting scheme could be used to improve performance.  However, it is worth noting that this limitation was previously identified by Heckemann et al. \cite{Heckemann2006a} and, subsequently, other groups have not even considered the case where an even number of candidate labels are considered (eg: \cite{Leung2010}).
